Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-09-01 papers/paper/ms.tex
update ms

%%%%%%%%%%%%%%%%
% INSTITUTIONS %
%%%%%%%%%%%%%%%%
\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}
\newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics,
77 Massachusetts Avenue, Cambridge, MA 02139, USA}
\newcommand{\berkeley}{Astronomy Department, University of California,
Berkeley, CA 94720, USA}
\newcommand{\cmu}{Department of Statistics \& Data Science, Carnegie
Mellon University, Pittsburgh, PA 15213, USA}
\newcommand{\osaka}{Department of Earth and Space Science, Osaka University, Osaka 560-0043, Japan}

@@

\received{2024 June 21}
\revised{---}
\accepted{---}
\shorttitle{Kepler's Demographic Cliff}

@@

\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}

@@

\begin{figure}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.45\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}

		\vspace{-0.3cm}
		\subfloat{
			\includegraphics[width=0.45\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
  \caption{{\bf The stars.}  Our analysis focuses on stars observed by
  Kepler with previously reported rotation periods (gray points).  The
  rotation periods are primarily drawn from
  \citet{Santos_2019,Santos_2021}.  About half of the stars with
  rotation periods are suitable for gyrochronology (dark gray points),
  based on factors including their non-binarity and proximity to the
  main sequence (orange line in lower panel; see
  Section~\ref{subsec:flags}).  Some host ``confirmed'' or
  ``candidate'' KOIs that meet additional planetary quality criteria
  (yellow points; Section~\ref{subsec:plflags}).  Surface gravities
  and effective temperatures were derived photometrically by
  \citet{Berger_2020a_catalog}.  Isochrones in the lower panel are
  from MIST \citep{Choi_2016}.
	}
	\label{fig:stellarprops}
\end{figure}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  The gray lines are \replaced{``mean fits'' to the
    rotation sequences of open clusters}{seventh-order polynomials
    fitted to the slow rotation sequences of open clusters \citep[see Section~3.3 of][]{Bouma_2023}}.  The 2-D histogram of the
    stellar sample (left) includes only apparently single stars near
    the main sequence with $\log g$$>$4.2, RUWE$<$1.4, and
    temperatures of 3800--6200\,K.  The planet hosts (right) require
    the same stellar cuts, and include only the confirmed and
    candidate planets described in Section~\ref{subsec:plflags}.  The
    total number of stars $N_{\rm s}$ and planets $N_{\rm p}$ are noted in
    the relevant panels.  A simple estimate for the number of young
    planets discovered by Kepler follows by counting points below the
    \deleted{mean }rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

Finally, since our scope is focused on rotation-based ages, we
restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.  The slowest-rotating FGK stars in the open clusters
used to calibrate our gyrochronology model have $P_{\rm
rot}$$\approx$35\,days.  Figure~\ref{fig:prot_vs_teff} shows a subset
of the resulting \nuniqstarsantosrot\ Kepler stars with rotation
periods compiled from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our extended KOI list.\added{  The gray
lines in this figure are the polynomials described by \citet{Bouma_2023}
that 
fit the slow rotation sequences of the Pleiades \citep{Rebull_2016},
Praesepe \citep{Rampalli_2021}, NGC-6811 \citep{Curtis_2019_ngc6811}, NGC-6819 \citep{Meibom_2015},
Ruprecht-147 \citep{Curtis_2020}, and M67 \citep{Barnes2016,Dungee_2022,Gruner_2023}.}

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.25cm}
	\caption{{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected from 2009--2024.  Lines are the ``mean''
    isochrones from \citet{Jeffries_2023}.  The intrinsic dispersion
    around these isochrones becomes much larger than changes in the
    mean at $\gtrsim$1\,Gyr (see Figure~\ref{fig:models}).  Some stars
    with lithium detections do not have detected rotation periods, and
    vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

We began this work with two questions: how wrong is the assumption of
a uniform age distribution for stars in the galactic thin disk?  And
why are only $\approx$50 sub-gigayear transiting planets known, rather
than the $\approx$500 that would be expected under the assumption of a
uniform star formation \replaced{history}{rate} for the $\approx$5{,}000 known planets?

@@

\added{Significant care is 
needed when assessing not only the reliability of rotational ages, but also any implications of measured age distributions.
Despite a slew of stellar quality flags, including on metallicity, evolutionary state,
and binarity, we expect that up to a third of the ``apparently clean'' 
Kepler stars with $P_{\rm rot}$$<$10\,days
could potentially be influenced by unresolved binaries (Figure~\ref{fig:hist_tgyro_corr}).
A separate nuance is in understanding that the observed age distribution does not 
directly map onto a star formation history: the kinematic heating of stars in the
disk in particular produces a major bias in the age distribution of stars, and therefore of planets,
observed by pencil-beam surveys such as Kepler.}

>>>>

%%%%%%%%%%%%%%%%
% INSTITUTIONS %
%%%%%%%%%%%%%%%%
\newcommand{\caltech}{Department of Astronomy, California Institute of Technology, Pasadena, CA 91125, USA}
\newcommand{\carnegie}{Observatories of the Carnegie Institution for Science, Pasadena, CA 91101, USA}
\newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics,
77 Massachusetts Avenue, Cambridge, MA 02139, USA}
\newcommand{\berkeley}{Astronomy Department, University of California,
Berkeley, CA 94720, USA}
\newcommand{\cmu}{Department of Statistics \& Data Science, Carnegie
Mellon University, Pittsburgh, PA 15213, USA}
\newcommand{\osaka}{Department of Earth and Space Science, Osaka University, Osaka 560-0043, Japan}

@@

\received{2024 June 21}
\revised{2024 September 5}
\accepted{---}
\shorttitle{Kepler's Demographic Cliff}

@@

\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}
\affiliation{\carnegie}

@@

\begin{figure}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.42\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}

		\vspace{-0.3cm}
		\subfloat{
			\includegraphics[width=0.42\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
  \caption{{\bf The stars.}  Our analysis focuses on stars observed by
  Kepler with previously reported rotation periods (gray points).  The
  rotation periods are primarily drawn from
  \citet{Santos_2019,Santos_2021}.  About half of the stars with
  rotation periods are suitable for gyrochronology (dark gray points),
  based on factors including their non-binarity and proximity to the
  main sequence (orange line in \replaced{lower}{upper} panel; see
  Section~\ref{subsec:flags}).  Some host ``confirmed'' or
  ``candidate'' KOIs that meet additional planetary quality criteria
  (yellow points; Section~\ref{subsec:plflags}).  Surface gravities
  and effective temperatures were derived photometrically by
  \citet{Berger_2020a_catalog}.  Isochrones in the \replaced{lower}{upper} panel are
  from MIST \citep{Choi_2016}.
	}
	\label{fig:stellarprops}
\end{figure}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.46\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.46\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  The gray lines are \replaced{``mean fits'' to the
    rotation sequences of open clusters}{polynomials
    fitted to the slow rotation sequences of open clusters}.  The 2-D histogram of the
    stellar sample (left) shows only apparently single stars near
    the main sequence with $\log g$$>$4.2, RUWE$<$1.4, and
    temperatures of 3800--6200\,K.  The planet hosts (right) require
    the same stellar cuts, and include only the confirmed and
    candidate planets described in Section~\ref{subsec:plflags}.  The
    total number of stars $N_{\rm s}$ and planets $N_{\rm p}$ are noted in
    the relevant panels.  A simple estimate for the number of young
    planets discovered by Kepler follows by counting points below the
    \deleted{mean }rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

Finally, since our scope is focused on rotation-based ages, we
restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.  The slowest-rotating FGK stars in the open clusters
used to calibrate our gyrochronology model have $P_{\rm
rot}$$\approx$35\,days.  Figure~\ref{fig:prot_vs_teff} shows a subset
of the resulting \nuniqstarsantosrot\ Kepler stars with rotation
periods compiled from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our extended KOI list.\added{  The gray
lines in this figure are the seventh-order polynomials described by \citet{Bouma_2023}
that 
fit the slow rotation sequences of the Pleiades \citep{Rebull_2016},
Praesepe \citep{Rampalli_2021}, NGC-6811 \citep{Curtis_2019_ngc6811}, NGC-6819 \citep{Meibom_2015},
Ruprecht-147 \citep{Curtis_2020}, and M67 \citep{Barnes2016,Dungee_2022,Gruner_2023}.}

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.5cm}
	\caption{{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected from 2009--2024.  Lines are the ``mean''
    isochrones from \citet{Jeffries_2023}.  The intrinsic dispersion
    around these isochrones becomes much larger than changes in the
    mean at $\gtrsim$1\,Gyr (see Figure~\ref{fig:models}).  Some stars
    with lithium detections do not have detected rotation periods, and
    vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

We began \deleted{this work }with two questions: how wrong is the assumption of
a uniform age distribution for stars in the galactic thin disk?  And
why are only $\approx$50 sub-gigayear transiting planets known, rather
than the $\approx$500 that would be expected under the assumption of a
uniform star formation \replaced{history}{rate} for the $\approx$5{,}000 known planets?

@@

\added{Significant care is 
needed not only when assessing the reliability of individual rotational ages, but also when interpreting distributions of rotational ages inferred from many stars.
Despite a slew of stellar quality flags, including on metallicity, evolutionary state,
and binarity, we expect that up to a third of our ``apparently clean'' 
Kepler stars with $P_{\rm rot}$$<$10\,days
could potentially be influenced by unresolved binaries (Figure~\ref{fig:hist_tgyro_corr}).
An additional nuance is that even after correcting for effects
that can bias rotational ages, our observed age distribution does not 
directly yield a star formation history: 
particularly for pencil-beam surveys such as Kepler,
kinematic heating of stars in the
thin disk significantly sculpts the stellar (and therefore planetary) age distribution.}
